---
title: "Intro to Visualisations in R"
tutorial:
  id: "yousaydata.introvis"
  version: 3.0.0
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: simplex
    includes:
      before_body: "./www/branding.html"
runtime: shiny_prerendered
description: >
  Learn some basic concepts of ggplot2.
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(tidytuesdayR)
library(skimr)

rolling_stone <- tt_load(2024, week = 19)[["rolling_stone"]]
gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)
```


## 1.1 The library we use

One of the most central libraries in modern R is `tidyverse`. We are going to use it throughout. `tidyverse` is essentially a collection of many different libraries for working with rectangular data. We are also using the library `tidytuesdayR` to access the some example datasets.

```{r echo=TRUE, eval=FALSE}
library(tidyverse)
library(tidytuesdayR)
library(skimr)

rolling_stone <- tt_load(2024, week = 19)[["rolling_stone"]]
```

In what follows let's try to answer whether newer albums have a higher spotify popularity. First we can have a look at it:

```{r lookup, exercise=TRUE}
rolling_stone
```

That just shows us the dataset itself. If we want a bit more information we can use `skim` from the `skimr` package:

```{r lookup2, exercise=TRUE}
skim(rolling_stone)
```

But often it is better to visualise the data. We can use `ggplot` for that. Here is a simple scatter plot:

```{r firstplot, exercise=TRUE}
ggplot(data = rolling_stone) +
  geom_point(mapping = aes(x = release_year, y = spotify_popularity))
```

So we see that `ggplot` has the following simple structure:

```{r echo=T}
# ggplot(data = <DATA>) +
# <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

## 1.2 Exercises

1. Run `ggplot(data = rolling_stone)`. What do you see?

```{r ex121, exercise=TRUE}

```

2. How many rows are in `rolling_stone`? How many columns?

```{r ex122, exercise=TRUE}

```

3. Make a scatterplot of `debut_album_release_year` vs `weeks_on_billboard` (replace the `<...>` in the code).

```{r ex123, exercise=TRUE}
ggplot(data = <DATA>) +
<GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

4. What happens if you make a scatterplot of `type` vs `artist_gender`? Why is the plot not very useful?

```{r ex125, exercise=TRUE}
ggplot(data = <DATA>) +
<GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```


## 2.1 Aesthetics 

The grammar of graphics attempts to translate features of the data into visual features of the plot. The `aes` function is central to this. Aside from the obvious x and y values, we can also map other variables to different aesthetics. 

For instance the colour.

```{r colmark, exercise=T}
ggplot(data = rolling_stone) + 
  geom_point(mapping = aes(x = release_year, y = spotify_popularity, color = artist_gender))
```

Of course there are more aesthetics and options.

Like size:
```{r sizemark, exercise=T}
ggplot(data = rolling_stone) + 
  geom_point(mapping = aes(x = release_year, y = spotify_popularity, size = weeks_on_billboard))
```

Like the alpha value of a colour:
```{r alphamark, exercise=T}
ggplot(data = rolling_stone) + 
  geom_point(mapping = aes(x = release_year, y = spotify_popularity, alpha = weeks_on_billboard))
```

Or shape:
```{r shapemark, exercise=T}
ggplot(data = rolling_stone) + 
  geom_point(mapping = aes(x = release_year, y = spotify_popularity, shape = artist_gender))
```

If I want to make a general change, I am doing this outside of the `aes` function:
```{r allmark, exercise=T}
ggplot(data = rolling_stone) + 
  geom_point(mapping = aes(x = release_year, y = spotify_popularity), size = 2)
```

## 2.2. Exercises

1. What’s gone wrong with this code? Why are the points not blue? Can you correct it?

```{r ex221, exercise=T}
ggplot(data = rolling_stone) + 
  geom_point(mapping = aes(x = release_year, y = spotify_popularity, color = "blue"))
```

2. Which variables in `rolling_stone` are categorical? Which variables are continuous? How can you see this information when you run `rolling_stone`?

```{r ex222, exercise=T}

```

3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

```{r ex223, exercise=T}

```

4. What happens if you map the same variable to multiple aesthetics?

```{r ex224, exercise=T}

```

5. What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)

```{r ex225, exercise=T}

```

6. What happens if you map an aesthetic to something other than a variable name, like aes(colour = years_between > 20)? Note, you’ll also need to specify x and y.

```{r ex226, exercise=T}

```

## 2.3. Facets

We can also use facets for our visualisations. Either we wrap it by one variable with `facet_wrap`:

```{r echo=T}
ggplot(data = rolling_stone) + 
  geom_point(mapping = aes(x = release_year, y = spotify_popularity))  + 
  facet_wrap(~ artist_gender, ncol = 2)
```

Or produce a two-dimensional grid with `facet_grid`:

```{r echo=T}
ggplot(data = rolling_stone) + 
  geom_point(mapping = aes(x = release_year, y = spotify_popularity)) + 
  facet_grid(artist_gender ~ type)
```

Feel free to try it yourself:

```{r ptt23, exercise = T}

```

## 3.1 GEOMs

I am sure you do not want to just produce scatter plots all the time. `ggplot` has you covered for all kinds of plots:

What's the difference here?

```{r echo=T}
ggplot(data = rolling_stone) + 
  geom_point(mapping = aes(x = release_year, y = spotify_popularity))

ggplot(data = rolling_stone) + 
  geom_smooth(mapping = aes(x = release_year, y = spotify_popularity))
```

With other types of plots, `ggplot` gives you access to another set of aesthetics. For instance, the `linetype` aesthetic:

```{r echo=T}
ggplot(data = rolling_stone) + 
  geom_smooth(mapping = aes(x = release_year, y = spotify_popularity, linetype = artist_gender))
```

In addition to switching `geom_`-functions you can also visually group the data or modify the legend:

```{r ptt31, exercise = T}
# ggplot(data = rolling_stone) + 
#   geom_smooth(mapping = aes(x = release_year, y = spotify_popularity, group = artist_gender))

# ggplot(data = rolling_stone) + 
#   geom_smooth(mapping = aes(x = release_year, y = spotify_popularity, col = artist_gender), show.legend = F)
```

It is also possible to use multiple different `geom`-functions. You essentially stack the visualisations on top of each other. `ggplot` works in layers!

```{r, echo=T}
ggplot(data = rolling_stone) + 
  geom_point(mapping = aes(x = release_year, y = spotify_popularity)) +
  geom_smooth(mapping = aes(x = release_year, y = spotify_popularity, col = artist_gender))
```

Try to avoid code duplication, though. The `geom`-functions can inhere from their parent `ggplot`-function. Check it out yourself: 

```{r ptt31a, exercise=T}
ggplot(data = rolling_stone, mapping = aes(x = release_year, y = spotify_popularity)) + 
  geom_point() +
  geom_smooth(mapping = aes(col = artist_gender))
```

If you do this, you can always overwrite inherited settings for individual layers. What do you think this will produce? 

```{r ptt31b, exercise=T}
ggplot(data = rolling_stone, mapping = aes(x = release_year, y = spotify_popularity)) + 
  geom_point(mapping = aes(col = artist_gender), show.legend = F) +
  geom_smooth()
```

## 3.2. Exercises

1. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

```{r ex321, exercise=T}

```

2. Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

```
ggplot(data = rolling_stone, mapping = aes(x = release_year, y = spotify_popularity)) + 
  geom_point(aes(color = type), show.legend = F) + 
  geom_smooth(aes(linetype = type), col = "black", se = FALSE)
``` 

```{r ex322, exercise=T}

```

3. What does the se argument to geom_smooth() do?

```{r ex324, exercise=T}

```