---
title: "Intro to Data Transformation in R"
tutorial:
  id: "yousaydata.datatramsform"
  version: 2.0.0
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: simplex
    includes:
      before_body: "./www/branding.html"
runtime: shiny_prerendered
description: >
  Learn some basic concepts of data transformation with R.
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)
```

## 1.1. Getting Started with `dplyr`

Once again we are going to work with one of the most central libraries in modern R: `tidyverse`. This time we work with the `dplyr` library. It's a library for working and cleaning up rectangular data.

If you start a fresh R session, you will have to load the tidyverse again.

```{r echo=TRUE}
library(tidyverse)
```

Let's also install a package that contains some example data:

```{r echo=TRUE, eval=FALSE}
install.packages("nycflights13")
```

And tell R that we want to use it:

```{r echo=TRUE}
library(nycflights13)
```

`nycflights13` just has a bunch of tables on which we can practice some data cleaning techniques. The tables are related to flights that were scheduled to depart from any of the three NYC airports in 2013. We are going to work with the `flights` library throughout this class:

```{r sc111, exercise=TRUE}
flights
```

### `dplyr` Functions

Since we are assuming we are working with tidy data, we will refer to the rows in the flight data set as **observations** and to the columns in the data set as **variables** in today's class. The main `dplyr` functions we are going to look at today are:

-   `filter()`: pick observations based on their values.
-   `arrange()`: reorder observations.
-   `select()`: pick variables based on their type or name.
-   `mutate()` / `transmute()`: create new variables with functions of existing variables.
-   `summarise()`: collapse many values down to a single summary.
-   `group_by()`: grouping your observations before doing any of it (especially with `mutate()` and `summarise()`)
-   `across()`: selecting several variables before doing any of it (especially with `mutate()` and `summarise()`)

## 1.2. `filter()`

Let's have a look at the flights table. If we want to know which flights left on Nov 8 2013, we could simply do:

```{r sc121, exercise=TRUE}
filter(flights, month == 11, day == 8)
```

And if we want it saved into our environment, we can **assign** it:

```{r sc122, exercise=TRUE}
(flights_nov_8 <- filter(flights, month == 11, day == 8))

```

*Does anyone know why I put brackets around the expression?*

Please note the `==`. A common mistake is to only use one `=`.

As you can see, `filter()` tests each value of selected variables with a condition. If this evaluates to `TRUE`, it keeps the observations, otherwise it drops it.

#### Caveat: Sometimes a computer evaluates things differently than us!

Think about the following expressions. Do they evaluate to true or false?

```{r sc123, exercise=TRUE}
sqrt(2) ^ 2 == 2
```

```{r sc124, exercise=TRUE}
1 / 49 * 49 == 1
```

Every computer has a build in precision, so they stop calculating and do not go on forever solving a simple equation. You can access the precision of your system like this:

```{r echo=TRUE}
.Machine$double.eps
```

If you work with very long floats it might be beneficial to use the `dplyr` function `near()` instead. `near()` takes the precision into account:

```{r echo=TRUE}
near(sqrt(2) ^ 2,  2)
```

### Combining Conditions

`filter()` can deal with all of R's boolean operators: !, &, \|, and `xor()`, but you have to be precise:

```{r sc125, exercise=TRUE}
# this works
filter(flights, month == 11 | month == 12)

# this does not
filter(flights, month == 11 | 12)

# this works again
nov_dec <- filter(flights, month %in% c(11, 12))
```

Sometimes you can give conditions in a positive or a negative way:

```{r sc126, exercise=TRUE}
# this produces the same results
x <- filter(flights, !(arr_delay > 120 | dep_delay > 120))
y <- filter(flights, arr_delay <= 120, dep_delay <= 120)

identical(x, y)
```

Really think about your conditions when you design more complicated queries. Especially when part of a condition evaluates to `NA` you might have a problem, because `NA`s are contagious:

```{r sc127, exercise=TRUE}
# missing values are contagious
NA > 5
10 == NA
NA + 10
NA / 2
NA == NA
```

The last one might be surprising, but let's think about it for a moment:

```{r sc128, exercise=TRUE}
# Let x be Mary's age. We don't know how old she is.
x <- NA

# Let y be John's age. We don't know how old he is.
y <- NA

# Are John and Mary the same age?
x == y
```

While this is sound, it can lead to bugs if we try things like `x == NA` (which would be wrong). If we need to find out whether a value is `NA` we can instead ask `is.na()`:

```{r echo=T}

df <- tibble(x = c(1, NA, 3))
filter(df, is.na(x) | x > 1)
```

## 1.3 Exercises

1.  Find all flights that:

    1.  Had an arrival delay of two or more hours

    ```{r ex131, exercise=TRUE}

    ```

    2.  Flew to Houston (IAH or HOU)

    ```{r ex132, exercise=TRUE}

    ```

    3.  Were operated by United, American, or Delta

    ```{r ex133, exercise=TRUE}

    ```

    4.  Departed in summer (July, August, and September)

    ```{r ex134, exercise=TRUE}

    ```

    5.  Arrived more than two hours late, but didn't leave late

    ```{r ex135, exercise=TRUE}

    ```

    6.  Were delayed by at least an hour, but made up over 30 minutes in flight

    ```{r ex136, exercise=TRUE}

    ```

    7.  Departed between midnight and 6am (inclusive)

    ```{r ex137, exercise=TRUE}

    ```

## 2.1 `arrange()`

With arrange we can sort our observations in ascending order:

```{r sc211, exercise=TRUE}
arrange(flights, carrier, year, month, day)
```

Or in descending order:

```{r sc212, exercise=TRUE}
arrange(flights, desc(dep_delay))
arrange(flights, -dep_delay)
```

## 2.2 `arrange()` exercises

1.  Sort flights to find the most delayed flights. Find the flights that left earliest.

```{r ex221, exercise=TRUE}

```

2.  Which flights travelled the longest? Which travelled the shortest?

```{r ex222, exercise=TRUE}

```

3.  Sort flights to find the fastest flights.

```{r ex223, exercise=TRUE}

```

4.  How could you use `arrange()` to sort all missing values to the start? (Hint: use `is.na()`)

```{r ex224, exercise=TRUE}

```

## 3.1 `select()`

You can select a variable by index:

```{r sc310, exercise=TRUE}
select(flights, 4, 5, 1)
```

You can select a variable by name:

```{r sc311, exercise=TRUE}
select(flights, dep_time, sched_dep_time, year)
```

You cam also use either:

```{r sc312, exercise=TRUE}
select(flights, all_of("year", "month", "day"))
```

Or:

```{r sc314, exercise=TRUE}
select(flights, any_of("year", "month", "day"))
```

You can also select a range:

```{r sc315, exercise=TRUE}
select(flights, year:day)
```

Or the inverse:

```{r sc316, exercise=TRUE}
select(flights, !year:day)
```

There are also `starts_with()`, `ends_with()`, `contains()`, `matches()`, and `num_range()`.

In addition there is `everything()`, which might not sound useful, but is great for reordering:

```{r sc317, exercise=TRUE}
select(flights, carrier, everything())
```

An alternative for reordering with `select()` is to use `relocate()`:

```{r sc318, exercise=TRUE}
relocate(flights, carrier, .before = year)
```

You can also rename columns when you are selecting them:

```{r sc319, exercise=TRUE}
select(flights, tail_num = tailnum, everything())
```

An alternative is the `rename()` function:

```{r sc3191, exercise=TRUE}
rename(flights, tail_num = tailnum)
```

## 3.2 `select()` exercises

1.  Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.

```{r ex321, exercise=TRUE}

```

## 4.1 `mutate()` / `transmute()`

Let's create a smaller tibble, so everything we do fits on our screen:

```{r echo=TRUE}
flights_sml <- select(flights, 
                      year:day, 
                      ends_with("delay"), 
                      distance, 
                      air_time
)
```

Now let's create new variables based on variables in the data set:

```{r sc411, exercise=TRUE}
mutate(flights_sml,
       gain = dep_delay - arr_delay,
       speed = distance / air_time * 60
)
```

You can even use a variable you have just created to create another one in the same call!

```{r sc412, exercise=TRUE}
mutate(flights_sml,
       gain = dep_delay - arr_delay,
       hours = air_time / 60,
       gain_per_hour = gain / hours
)
```

If you only want to keep the columns you created use `transmute()`

```{r sc413, exercise=TRUE}
transmute(flights,
          gain = dep_delay - arr_delay,
          hours = air_time / 60,
          gain_per_hour = gain / hours
)
```

You can use all typical math functions, including integer division and modulo (remainder):

```{r sc414, exercise=TRUE}
transmute(flights,
          dep_time,
          hour = dep_time %/% 100,
          minute = dep_time %% 100
)
```

## 4.2 `mutate()` / `transmute()` exercises

1.  Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they're not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.

```{r ex421, exercise=TRUE}

```

2.  Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().

```{r ex422, exercise=TRUE}

```

## 5.1 `summarise()`

```{r sc511, exercise=TRUE}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```

## 6.1 `group_by()`

```{r sc611, exercise=TRUE}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

```{r sc612, exercise=TRUE}
flights  |>
  group_by(year, month, day) |>
  summarise(delay = mean(dep_delay, na.rm = TRUE))
```

```{r sc613, exercise=TRUE}
flights  |>
  group_by(year, month, day) |>
  summarise(delay = mean(dep_delay, na.rm = TRUE, .groups = "drop"))
```

```{r sc614, exercise=TRUE}
flights |> 
  group_by(origin, carrier) |> 
  summarise(count = n()) |> 
  mutate(prop_per_airport = count / sum(count)) |>
  arrange(-prop_per_airport)
```

## 7.1 `across()`

```{r sc711, exercise=TRUE}
flights  |>
  group_by(origin) |>
  summarise(
    across(
      c(dep_delay, arr_delay),
      .fns = list(mean, min, max),
      na.rm = T),
    scheduled_flights = n(),
    cancelled = sum(is.na(dep_time))
    ) |>
  mutate(
    prop_cancelled = cancelled / scheduled_flights)

```
